---
title: "California Proposition 50: Gerrymandering Analysis"
format: dashboard
---

```{r}
#| include: false
library(tidyverse)
library(scales)
library(plotly)

# Load all data and calculations
source_data <- function() {
  # Load cleaned data
  precinct_data <- read_csv("data/g24_sov_by_g24_svprec_clean.csv", show_col_types = FALSE)
  candidates <- read_csv("data/g24-candidates-by-district.csv", show_col_types = FALSE)
  ab604_results <- read_csv("data/ab604_district_results.csv", show_col_types = FALSE)

  # Helper functions
  calculate_wasted_votes <- function(votes_a, votes_b) {
    total_votes <- votes_a + votes_b
    win_threshold <- total_votes / 2
    wasted_a <- ifelse(votes_a > votes_b, votes_a - win_threshold, votes_a)
    wasted_b <- ifelse(votes_b > votes_a, votes_b - win_threshold, votes_b)
    return(list(wasted_a = wasted_a, wasted_b = wasted_b))
  }

  calculate_mean_median <- function(vote_shares) {
    mean(vote_shares, na.rm = TRUE) - median(vote_shares, na.rm = TRUE)
  }

  calculate_efficiency_gap <- function(votes_a, votes_b) {
    wasted <- calculate_wasted_votes(votes_a, votes_b)
    (sum(wasted$wasted_a) - sum(wasted$wasted_b)) / sum(votes_a + votes_b)
  }

  # Party mapping
  cong_candidates <- candidates %>% filter(DISTRICT_TYPE == "CNG")
  party_mapping <- cong_candidates %>%
    select(FIELD, DISTRICT) %>%
    mutate(party = case_when(
      grepl("DEM", FIELD) ~ "Democrat",
      grepl("REP", FIELD) ~ "Republican",
      TRUE ~ "Other"
    ))

  # Process 2024
  cong_vote_cols <- names(precinct_data)[grep("^CNG", names(precinct_data))]
  district_votes_2024 <- precinct_data %>%
    filter(!is.na(CDDIST)) %>%
    group_by(CDDIST) %>%
    summarise(across(all_of(cong_vote_cols), ~sum(., na.rm = TRUE)), .groups = "drop")

  party_totals_2024 <- district_votes_2024 %>%
    pivot_longer(cols = all_of(cong_vote_cols), names_to = "FIELD", values_to = "votes") %>%
    left_join(party_mapping, by = "FIELD", relationship = "many-to-many") %>%
    filter(votes > 0) %>%
    group_by(CDDIST, party) %>%
    summarise(total_votes = sum(votes), .groups = "drop") %>%
    pivot_wider(names_from = party, values_from = total_votes, values_fill = 0)

  dem_rep_2024 <- party_totals_2024 %>%
    select(CDDIST, Democrat, Republican) %>%
    mutate(
      total = Democrat + Republican,
      dem_share = Democrat / total,
      winner = if_else(Democrat > Republican, "D", "R")
    )

  # Process AB 604
  party_totals_ab604 <- ab604_results %>%
    pivot_longer(cols = all_of(cong_vote_cols), names_to = "FIELD", values_to = "votes") %>%
    left_join(party_mapping, by = "FIELD", relationship = "many-to-many") %>%
    filter(votes > 0) %>%
    group_by(AB604_DISTRICT, party) %>%
    summarise(total_votes = sum(votes), .groups = "drop") %>%
    pivot_wider(names_from = party, values_from = total_votes, values_fill = 0)

  dem_rep_ab604 <- party_totals_ab604 %>%
    select(AB604_DISTRICT, Democrat, Republican) %>%
    mutate(
      total = Democrat + Republican,
      dem_share = Democrat / total,
      winner = if_else(Democrat > Republican, "D", "R")
    )

  # Calculate metrics
  list(
    dem_rep_2024 = dem_rep_2024,
    dem_rep_ab604 = dem_rep_ab604,
    seats_2024_D = sum(dem_rep_2024$winner == "D"),
    seats_2024_R = sum(dem_rep_2024$winner == "R"),
    seats_ab604_D = sum(dem_rep_ab604$winner == "D"),
    seats_ab604_R = sum(dem_rep_ab604$winner == "R"),
    mm_2024 = calculate_mean_median(dem_rep_2024$dem_share),
    mm_ab604 = calculate_mean_median(dem_rep_ab604$dem_share),
    eg_2024 = calculate_efficiency_gap(dem_rep_2024$Democrat, dem_rep_2024$Republican),
    eg_ab604 = calculate_efficiency_gap(dem_rep_ab604$Democrat, dem_rep_ab604$Republican)
  )
}

data <- source_data()
```

# Overview {orientation="columns"}

## Column {width="50%"}

### 2024 District Map

```{r}
#| title: "Seat Distribution (2024)"
tibble(
  Party = c("Democrats", "Republicans"),
  Seats = c(data$seats_2024_D, data$seats_2024_R)
) %>%
  plot_ly(labels = ~Party, values = ~Seats, type = "pie",
          marker = list(colors = c("#2E5DA6", "#C92C34")),
          textinfo = "label+value+percent") %>%
  layout(showlegend = TRUE)
```

### Key Metrics (2024)

```{r}
#| content: valuebox
#| title: "Democratic Seats"
#| icon: "people-fill"
#| color: primary
list(
  value = data$seats_2024_D,
  caption = paste0(round(100 * data$seats_2024_D / 53, 1), "% of 53 seats")
)
```

```{r}
#| content: valuebox
#| title: "Mean-Median Score"
#| icon: "graph-up"
list(
  value = round(data$mm_2024, 4),
  caption = "Closer to 0 = less gerrymandering"
)
```

```{r}
#| content: valuebox
#| title: "Efficiency Gap"
#| icon: "bar-chart"
list(
  value = round(data$eg_2024, 4),
  caption = "Negative = Republicans waste more"
)
```

## Column {width="50%"}

### AB 604 Proposed Map

```{r}
#| title: "Seat Distribution (AB 604)"
tibble(
  Party = c("Democrats", "Republicans"),
  Seats = c(data$seats_ab604_D, data$seats_ab604_R)
) %>%
  plot_ly(labels = ~Party, values = ~Seats, type = "pie",
          marker = list(colors = c("#2E5DA6", "#C92C34")),
          textinfo = "label+value+percent") %>%
  layout(showlegend = TRUE)
```

### Key Metrics (AB 604)

```{r}
#| content: valuebox
#| title: "Democratic Seats"
#| icon: "people-fill"
#| color: primary
list(
  value = data$seats_ab604_D,
  caption = paste0(round(100 * data$seats_ab604_D / 52, 1), "% of 52 seats")
)
```

```{r}
#| content: valuebox
#| title: "Mean-Median Score"
#| icon: "graph-up"
list(
  value = round(data$mm_ab604, 4),
  caption = "Closer to 0 = less gerrymandering"
)
```

```{r}
#| content: valuebox
#| title: "Efficiency Gap"
#| icon: "bar-chart"
list(
  value = round(data$eg_ab604, 4),
  caption = "Negative = Republicans waste more"
)
```

# Analysis {orientation="rows"}

## Row {height="30%"}

### Change in Seats

```{r}
#| content: valuebox
#| title: "Democratic Seat Change"
#| icon: "arrow-up-right"
#| color: primary
seat_change <- data$seats_ab604_D - data$seats_2024_D
list(
  value = paste0(if_else(seat_change >= 0, "+", ""), seat_change),
  caption = paste0("From ", data$seats_2024_D, " to ", data$seats_ab604_D, " seats")
)
```

### Change in Mean-Median

```{r}
#| content: valuebox
#| title: "Mean-Median Change"
#| icon: "graph-up-arrow"
mm_change <- data$mm_ab604 - data$mm_2024
list(
  value = sprintf("%+.4f", mm_change),
  caption = if_else(abs(mm_change) < 0.01, "Minimal change",
                    "Notable change in symmetry")
)
```

### Change in Efficiency Gap

```{r}
#| content: valuebox
#| title: "Efficiency Gap Change"
#| icon: "bar-chart-steps"
eg_change <- data$eg_ab604 - data$eg_2024
list(
  value = sprintf("%+.4f", eg_change),
  caption = if_else(abs(eg_change) < 0.05, "Modest change",
                    "Significant change in efficiency")
)
```

## Row {height="70%"}

### Vote Share Distributions

```{r}
#| title: "Democratic Vote Share by District"
comparison_data <- bind_rows(
  data$dem_rep_2024 %>% mutate(Map = "2024", District = as.character(CDDIST)) %>% select(-CDDIST),
  data$dem_rep_ab604 %>% mutate(Map = "AB 604", District = as.character(AB604_DISTRICT)) %>% select(-AB604_DISTRICT)
)

plot_ly(comparison_data, x = ~dem_share, color = ~Map, type = "histogram",
        colors = c("2024" = "gray", "AB 604" = "steelblue"),
        nbinsx = 20, alpha = 0.6) %>%
  layout(
    barmode = "overlay",
    xaxis = list(title = "Democratic Vote Share", tickformat = ".0%"),
    yaxis = list(title = "Number of Districts"),
    title = "Distribution of Democratic Vote Share: 2024 vs AB 604"
  )
```

### District Competitiveness

```{r}
#| title: "Competitiveness by Map"
comp_summary <- comparison_data %>%
  mutate(competitiveness = case_when(
    abs(dem_share - 0.5) < 0.05 ~ "Highly Competitive (<5% margin)",
    abs(dem_share - 0.5) < 0.10 ~ "Competitive (<10% margin)",
    abs(dem_share - 0.5) < 0.20 ~ "Leaning (<20% margin)",
    TRUE ~ "Safe (>20% margin)"
  )) %>%
  count(Map, competitiveness)

plot_ly(comp_summary, x = ~Map, y = ~n, color = ~competitiveness, type = "bar") %>%
  layout(
    barmode = "stack",
    xaxis = list(title = ""),
    yaxis = list(title = "Number of Districts"),
    title = "District Competitiveness Comparison"
  )
```

# Methodology

## Column

### Data Sources

This analysis uses official election data from the California Secretary of State and the Statewide Database:

- **2024 General Election Results**: Precinct-level statement of vote (SV and SR precincts)
- **AB 604 District Boundaries**: Census block assignments from the Assembly Elections Committee
- **Candidate Information**: Party affiliations for all congressional candidates
- **Census Block Data**: Registration precinct to block correspondence from Statewide Database

All data accessed October-November 2025.

### Data Cleaning Process

**Part 2: Cleaning 2024 Election Data**

1. Loaded 51,123 precinct-level records from 2024 General Election
2. Handled suppressed values (marked as "***" for privacy) by converting to NA
3. Converted vote columns to numeric type
4. Validated precinct key uniqueness and district assignments
5. Aggregated votes to 53 congressional districts

**Part 5: Simulating AB 604 Results (Proportional Allocation)**

1. Loaded SR precinct voting data (geographic precincts with stable boundaries)
2. Obtained census block-to-precinct mapping with voter registration counts
3. Loaded AB 604 block assignments (519,723 census blocks mapped to 52 new districts)
4. Allocated SR precinct votes to census blocks proportionally based on registered voters
5. Aggregated census block votes to AB 604 districts

This approach assumes voters are distributed proportionally to registered voters within each precinct.

### Gerrymandering Metrics

**Mean-Median Score**

Measures asymmetry in vote distribution. Calculated as the difference between mean and median vote share across districts for one party.

- **Positive score**: The favored party's votes are "packed" into fewer districts
- **Negative score**: The opposing party's votes are packed
- **Near zero**: More symmetric distribution

**Efficiency Gap**

Measures wasted votes - votes that don't contribute to winning a seat. Wasted votes include:

- All votes for losing candidates
- Surplus votes (above 50%) for winning candidates

Formula: (Wasted votes for party A - Wasted votes for party B) / Total votes

- **Positive gap**: Party A wastes more votes (disadvantaged)
- **Negative gap**: Party B wastes more votes (disadvantaged)
- **Near zero**: Votes wasted equally

### Caveats and Limitations

1. **Single Election Analysis**: These metrics use only 2024 election results. Gerrymandering assessment is more robust across multiple election cycles.

2. **Third-Party Candidates**: Our analysis focuses on Democrat vs Republican races. Districts with significant third-party or independent candidates may have different dynamics.

3. **Proportional Allocation Assumption**: AB 604 analysis assumes voters are distributed within precincts proportionally to registered voters. Actual turnout patterns may differ.

4. **Block Assignment Timing**: Census blocks may have changed between 2020 census and 2024 election, introducing minor geographic discrepancies.

5. **No Geographic Context**: These metrics don't account for Communities of Interest, geographic constraints, or Voting Rights Act requirements that may justify certain district shapes.

6. **Statistical Thresholds**: There is no universally agreed threshold for "acceptable" levels of these metrics. Courts have used various standards.

### Research Question

**Will the adoption of the new district map (AB 604) lead to an increase in partisan advantage relative to 2024?**

**Answer**: Based on our analysis:

- **Seat Distribution**: AB 604 would result in `r data$seats_ab604_D - data$seats_2024_D` additional Democratic seats
- **Mean-Median Score**: Changed from `r round(data$mm_2024, 4)` to `r round(data$mm_ab604, 4)` (change: `r sprintf("%+.4f", data$mm_ab604 - data$mm_2024)`)
- **Efficiency Gap**: Changed from `r round(data$eg_2024, 4)` to `r round(data$eg_ab604, 4)` (change: `r sprintf("%+.4f", data$eg_ab604 - data$eg_2024)`)

The direction and magnitude of these changes suggest `r if_else(abs(data$mm_ab604) > abs(data$mm_2024) | abs(data$eg_ab604) > abs(data$eg_2024), "AB 604 may increase partisan advantage", "AB 604 may not substantially increase partisan advantage")`.

---

*Analysis conducted for STAT 133, Fall 2025*

*Data sources: California Secretary of State, Statewide Database, Assembly Elections Committee*
