---
title: "Gerrymandering Metrics"
format: typst
---

# Gerrymandering Analysis: 2024 vs AB 604

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(scales)
```

## Helper Functions

```{r}
# Calculate wasted votes for two parties
calculate_wasted_votes <- function(votes_a, votes_b) {
  # votes_a and votes_b are vectors of votes by district
  total_votes <- votes_a + votes_b
  win_threshold <- total_votes / 2

  # For party A:
  # - In districts A wins: wasted = votes above threshold (surplus)
  # - In districts A loses: wasted = all votes
  wasted_a <- ifelse(votes_a > votes_b,
                     votes_a - win_threshold,  # surplus in won districts
                     votes_a)                   # all votes in lost districts

  # For party B: same logic
  wasted_b <- ifelse(votes_b > votes_a,
                     votes_b - win_threshold,
                     votes_b)

  return(list(wasted_a = wasted_a, wasted_b = wasted_b))
}

# Calculate mean-median score
calculate_mean_median <- function(vote_shares) {
  # vote_shares: vector of vote shares for one party across districts
  mean_share <- mean(vote_shares, na.rm = TRUE)
  median_share <- median(vote_shares, na.rm = TRUE)
  return(mean_share - median_share)
}

# Calculate efficiency gap
calculate_efficiency_gap <- function(votes_a, votes_b) {
  wasted <- calculate_wasted_votes(votes_a, votes_b)
  total_votes <- sum(votes_a + votes_b)
  gap <- (sum(wasted$wasted_a) - sum(wasted$wasted_b)) / total_votes
  return(gap)
}

cat("✓ Helper functions loaded\n")
```

# Part 4: 2024 Election Results and the 2024 District Map

## Load and Prepare Data

```{r}
# Load cleaned precinct data
precinct_data <- read_csv("data/g24_sov_by_g24_svprec_clean.csv", show_col_types = FALSE)

# Load candidate information
candidates <- read_csv("data/g24-candidates-by-district.csv", show_col_types = FALSE)

# Get congressional candidates only
cong_candidates <- candidates %>%
  filter(DISTRICT_TYPE == "CNG")

cat("Data loaded:\n")
cat("  Precincts:", nrow(precinct_data), "\n")
cat("  Congressional candidates:", nrow(cong_candidates), "\n\n")
```

## Map Candidates to Parties

```{r}
# Create mapping of vote columns to parties
party_mapping <- cong_candidates %>%
  select(FIELD, DISTRICT, NAME) %>%
  mutate(party = case_when(
    grepl("DEM", FIELD) ~ "Democrat",
    grepl("REP", FIELD) ~ "Republican",
    grepl("IND", FIELD) ~ "Independent",
    grepl("LIB", FIELD) ~ "Libertarian",
    grepl("GRN", FIELD) ~ "Green",
    TRUE ~ "Other"
  ))

cat("Party breakdown:\n")
print(table(party_mapping$party))
cat("\n")
```

## Aggregate Votes by Party and District (2024)

```{r}
# Get congressional vote columns
cong_vote_cols <- names(precinct_data)[grep("^CNG", names(precinct_data))]

# Aggregate to district level
district_votes_2024 <- precinct_data %>%
  filter(!is.na(CDDIST)) %>%
  group_by(CDDIST) %>%
  summarise(across(all_of(cong_vote_cols), ~sum(., na.rm = TRUE)),
            .groups = "drop")

# Pivot to long format and join with party info
district_party_votes_2024 <- district_votes_2024 %>%
  pivot_longer(cols = all_of(cong_vote_cols),
               names_to = "FIELD",
               values_to = "votes") %>%
  left_join(party_mapping, by = "FIELD") %>%
  filter(votes > 0)

# Sum votes by district and party
party_totals_2024 <- district_party_votes_2024 %>%
  group_by(CDDIST, party) %>%
  summarise(total_votes = sum(votes), .groups = "drop") %>%
  pivot_wider(names_from = party, values_from = total_votes, values_fill = 0)

cat("2024 district-level results by party:\n")
print(head(party_totals_2024, 10))
cat("\n")
```

## Calculate Democrat vs Republican Metrics (2024)

```{r}
# For gerrymandering metrics, focus on D vs R (exclude third parties for now)
dem_rep_2024 <- party_totals_2024 %>%
  select(CDDIST, Democrat, Republican) %>%
  mutate(
    total = Democrat + Republican,
    dem_share = Democrat / total,
    rep_share = Republican / total,
    winner = if_else(Democrat > Republican, "D", "R"),
    margin = abs(dem_share - rep_share)
  )

cat("=== 2024 Election Results ===\n\n")

# Seat counts
seats_2024 <- table(dem_rep_2024$winner)
cat("Seat Distribution:\n")
cat("  Democrats:", seats_2024["D"], "seats\n")
cat("  Republicans:", seats_2024["R"], "seats\n\n")

# Mean-median score
mm_score_2024 <- calculate_mean_median(dem_rep_2024$dem_share)
cat("Mean-Median Score:", round(mm_score_2024, 4), "\n")
cat("  (Positive = Democrats packed/cracked, Negative = Republicans packed/cracked)\n\n")

# Efficiency gap
eg_2024 <- calculate_efficiency_gap(dem_rep_2024$Democrat, dem_rep_2024$Republican)
cat("Efficiency Gap:", round(eg_2024, 4), "\n")
cat("  (Positive = Democrats waste more, Negative = Republicans waste more)\n\n")

# Wasted votes breakdown
wasted_2024 <- calculate_wasted_votes(dem_rep_2024$Democrat, dem_rep_2024$Republican)
total_wasted_dem <- sum(wasted_2024$wasted_a)
total_wasted_rep <- sum(wasted_2024$wasted_b)

cat("Wasted Votes:\n")
cat("  Democrats:", comma(total_wasted_dem), "\n")
cat("  Republicans:", comma(total_wasted_rep), "\n")
cat("  Difference:", comma(total_wasted_dem - total_wasted_rep), "\n\n")
```

## Visualize 2024 Results

```{r}
# Vote share distribution
ggplot(dem_rep_2024, aes(x = dem_share)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(dem_rep_2024$dem_share),
             color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median(dem_rep_2024$dem_share),
             color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean(dem_rep_2024$dem_share), y = Inf,
           label = paste0("Mean: ", percent(mean(dem_rep_2024$dem_share), 0.1)),
           vjust = 1.5, hjust = 1.1, color = "red") +
  annotate("text", x = median(dem_rep_2024$dem_share), y = Inf,
           label = paste0("Median: ", percent(median(dem_rep_2024$dem_share), 0.1)),
           vjust = 3, hjust = -0.1, color = "blue") +
  scale_x_continuous(labels = percent) +
  labs(title = "Distribution of Democratic Vote Share by District (2024)",
       subtitle = paste0("Mean-Median Score: ", round(mm_score_2024, 4)),
       x = "Democratic Vote Share",
       y = "Number of Districts") +
  theme_minimal()
```

# Part 6: 2024 Election Results and the AB 604 District Map

## Load AB 604 Results

```{r}
# Load AB 604 district results (created in Part 5)
ab604_results <- read_csv("data/ab604_district_results.csv", show_col_types = FALSE)

cat("AB 604 results loaded:", nrow(ab604_results), "districts\n\n")
```

## Aggregate AB 604 Votes by Party

```{r}
# Same vote columns, but for AB 604 districts
ab604_party_votes <- ab604_results %>%
  pivot_longer(cols = all_of(cong_vote_cols),
               names_to = "FIELD",
               values_to = "votes") %>%
  left_join(party_mapping, by = "FIELD") %>%
  filter(votes > 0)

# Sum by district and party
party_totals_ab604 <- ab604_party_votes %>%
  group_by(AB604_DISTRICT, party) %>%
  summarise(total_votes = sum(votes), .groups = "drop") %>%
  pivot_wider(names_from = party, values_from = total_votes, values_fill = 0)

cat("AB 604 district-level results by party:\n")
print(head(party_totals_ab604, 10))
cat("\n")
```

## Calculate Democrat vs Republican Metrics (AB 604)

```{r}
# Focus on D vs R
dem_rep_ab604 <- party_totals_ab604 %>%
  select(AB604_DISTRICT, Democrat, Republican) %>%
  mutate(
    total = Democrat + Republican,
    dem_share = Democrat / total,
    rep_share = Republican / total,
    winner = if_else(Democrat > Republican, "D", "R"),
    margin = abs(dem_share - rep_share)
  )

cat("=== AB 604 Projected Results ===\n\n")

# Seat counts
seats_ab604 <- table(dem_rep_ab604$winner)
cat("Seat Distribution:\n")
cat("  Democrats:", seats_ab604["D"], "seats\n")
cat("  Republicans:", seats_ab604["R"], "seats\n\n")

# Mean-median score
mm_score_ab604 <- calculate_mean_median(dem_rep_ab604$dem_share)
cat("Mean-Median Score:", round(mm_score_ab604, 4), "\n")
cat("  (Positive = Democrats packed/cracked, Negative = Republicans packed/cracked)\n\n")

# Efficiency gap
eg_ab604 <- calculate_efficiency_gap(dem_rep_ab604$Democrat, dem_rep_ab604$Republican)
cat("Efficiency Gap:", round(eg_ab604, 4), "\n")
cat("  (Positive = Democrats waste more, Negative = Republicans waste more)\n\n")

# Wasted votes
wasted_ab604 <- calculate_wasted_votes(dem_rep_ab604$Democrat, dem_rep_ab604$Republican)
total_wasted_dem_ab604 <- sum(wasted_ab604$wasted_a)
total_wasted_rep_ab604 <- sum(wasted_ab604$wasted_b)

cat("Wasted Votes:\n")
cat("  Democrats:", comma(total_wasted_dem_ab604), "\n")
cat("  Republicans:", comma(total_wasted_rep_ab604), "\n")
cat("  Difference:", comma(total_wasted_dem_ab604 - total_wasted_rep_ab604), "\n\n")
```

## Visualize AB 604 Results

```{r}
# Vote share distribution
ggplot(dem_rep_ab604, aes(x = dem_share)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(dem_rep_ab604$dem_share),
             color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median(dem_rep_ab604$dem_share),
             color = "blue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = mean(dem_rep_ab604$dem_share), y = Inf,
           label = paste0("Mean: ", percent(mean(dem_rep_ab604$dem_share), 0.1)),
           vjust = 1.5, hjust = 1.1, color = "red") +
  annotate("text", x = median(dem_rep_ab604$dem_share), y = Inf,
           label = paste0("Median: ", percent(median(dem_rep_ab604$dem_share), 0.1)),
           vjust = 3, hjust = -0.1, color = "blue") +
  scale_x_continuous(labels = percent) +
  labs(title = "Distribution of Democratic Vote Share by District (AB 604)",
       subtitle = paste0("Mean-Median Score: ", round(mm_score_ab604, 4)),
       x = "Democratic Vote Share",
       y = "Number of Districts") +
  theme_minimal()
```

## Comparison: 2024 vs AB 604

```{r}
cat("\n========================================\n")
cat("COMPARISON: 2024 vs AB 604\n")
cat("========================================\n\n")

# Seat changes
cat("Seat Distribution:\n")
cat("                    2024    AB 604   Change\n")
cat("  Democrats:       ", sprintf("%4d", seats_2024["D"]),
    "    ", sprintf("%4d", seats_ab604["D"]),
    "    ", sprintf("%+3d", seats_ab604["D"] - seats_2024["D"]), "\n")
cat("  Republicans:     ", sprintf("%4d", seats_2024["R"]),
    "    ", sprintf("%4d", seats_ab604["R"]),
    "    ", sprintf("%+3d", seats_ab604["R"] - seats_2024["R"]), "\n\n")

# Metric changes
cat("Gerrymandering Metrics:\n")
cat("                         2024      AB 604    Change\n")
cat("  Mean-Median Score:   ", sprintf("%7.4f", mm_score_2024),
    "   ", sprintf("%7.4f", mm_score_ab604),
    "   ", sprintf("%+7.4f", mm_score_ab604 - mm_score_2024), "\n")
cat("  Efficiency Gap:      ", sprintf("%7.4f", eg_2024),
    "   ", sprintf("%7.4f", eg_ab604),
    "   ", sprintf("%+7.4f", eg_ab604 - eg_2024), "\n\n")

# Interpretation
cat("Interpretation:\n")
if (abs(mm_score_ab604) > abs(mm_score_2024)) {
  cat("  ⚠ Mean-median score INCREASED in magnitude\n")
  cat("    Suggests AB 604 may introduce more partisan advantage\n")
} else {
  cat("  ✓ Mean-median score DECREASED in magnitude\n")
  cat("    Suggests AB 604 may reduce partisan advantage\n")
}

if (abs(eg_ab604) > abs(eg_2024)) {
  cat("  ⚠ Efficiency gap INCREASED in magnitude\n")
  cat("    Suggests AB 604 may waste more votes unevenly\n")
} else {
  cat("  ✓ Efficiency gap DECREASED in magnitude\n")
  cat("    Suggests AB 604 may distribute votes more efficiently\n")
}

cat("\n")
```

## Side-by-Side Comparison

```{r}
# Combine data for comparison
comparison_data <- bind_rows(
  dem_rep_2024 %>% mutate(map = "2024", district = as.character(CDDIST)) %>% select(-CDDIST),
  dem_rep_ab604 %>% mutate(map = "AB 604", district = as.character(AB604_DISTRICT)) %>% select(-AB604_DISTRICT)
)

# Plot both distributions
ggplot(comparison_data, aes(x = dem_share, fill = map)) +
  geom_histogram(bins = 20, alpha = 0.6, position = "identity", color = "white") +
  geom_vline(data = data.frame(
    map = c("2024", "AB 604"),
    mean_val = c(mean(dem_rep_2024$dem_share), mean(dem_rep_ab604$dem_share))
  ), aes(xintercept = mean_val, color = map), linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("2024" = "gray50", "AB 604" = "steelblue")) +
  scale_color_manual(values = c("2024" = "gray20", "AB 604" = "blue")) +
  scale_x_continuous(labels = percent) +
  labs(title = "Democratic Vote Share Distribution: 2024 vs AB 604",
       subtitle = "Dashed lines show mean values",
       x = "Democratic Vote Share",
       y = "Number of Districts",
       fill = "District Map",
       color = "Mean") +
  theme_minimal()
```
